Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-06 paper/main.tex
more shortening

%SHORT
\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
\captionsetup[figure]{format=myformat}
\renewcommand{\captionlabelfont}{\bf}
%SHORT

@@

\iffull{We only show a small selections of the programs that we can handle
automatically here.}{}  In \pref{app:cat} is a list of more than $30$
classes of challenging programs that we can automatically analyze.
\pref{sec:exper} contains a more detailed comparison with other tools
for \iffull{automatic}{} bound derivation.

@@

To find derivations in the quantitative Hoare logic automatically, we
have to focus on bounds that have a certain shape.  The general form
of \emph{potential functions} (or assertions) that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is again a simplified program state as introduced in \pref{sec:sem},
$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$
We often write $xy$ to denote the index $(x,y)$.  The family
$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense) for
potential functions.  This allows us to uniquely represent any linear
potential function $\Phi$ as a \emph{quantitative annotation} $Q =
(q_i)_{i \in I}$, that is, a family of rational numbers where only a
finite number of elements are not zero.
%
For soundness, we require that potential functions always give
non-negative results.  This is acheived by restricting coefficients
in a quantitative annotation to be always non-negative.

@@

In addition to the quantitative annotations our automatic amortized
analysis needs to maintain a minimal logical state to justify certain
operations on quantitative annotations.  For example when analyzing
the code $x \gets x + y$, it is helpful to know the sign of $y$ to
determine which intervals will increase or decrease.  The knowledge
needed by our rules can often be inferred by local reasoning (i.e., in
basic blocks without recursion and loops) within usual theories
(e.g. Presburger arithmetic or bit vectors).

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
%
Our implementation uses conjunctions of linear inequalities as
logical state.  We never compute fixpoints and take $\top$
as pre- and postconditions for functions. It has proved to be
sufficient for the variety of examples we designed and found
in the literature.

@@

@@

\and \Rule{Q:Return}

@@

@@

@@

\and \Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}

@@

\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')

@@

\\ Q_1 \succeq_{\Gamma_1} Q_2 

@@

\end{mathpar}
\caption{Inference rules of the quantitative analysis.}
\label{fig:auto}
\end{figure*}

@@

The inference system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The inference rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
These judgements correspond to the logic judgements $\Delta; B; R
\vdash \htriple P S Q$ where each assertion splits in two parts, a
logical part $\Gamma$ and a quantitative part $Q$.  That means that
$(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
$(\Gamma_R, Q_R)$ is the postcondition for return statements, and
$(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
simple Hoare triple.  We leave out the function context $\Delta$ in
the presentation to avoid cluttering notations.  Instead we assume a
fixed function context $\Delta$ that is implicit on all judgements.
Function contexts are treated exactly as in the logic.  We call
precondition the pair $(\Gamma, Q)$ and postcondition the pair
$(\Gamma', Q')$.  The correspondence between judgements of the
automation and the ones of the quantitative logic is made formal by
our soundness proof of the automatic amortized analysis at the end of
this section.

@@

We describe the automatic amortized analysis for a subset of
expressions of C light.  Assignments must have the form \code{x \gets
  y} or \code{x \gets x \pm y}.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This achieved by using a series of \emph{cost-free
  assignments} that we do not result in additional cost in the
analysis.  Non-linear operations such as \code{x \gets z*y} are simply
handled by assigning zero potential to coefficients like $q_{xa}$ and
$q_{ax}$ that contain $x$.

@@

Most of the rules correspond exactly to the respective rules of the
quantitative Hoare logic.  For example, the rule {\sc Q:Skip} simply
reuses its postcondition as precondition.  This is sound since
evaluating a $\code{skip}$ statement never costs any resources and
does not change the program state.  Similarly, the rule {\sc Q:Tick}
enforces that the quantitative annotation of the precondition is at
least $M_t(n)$ to pay for the resource consumption.  Since the program
state is also unchanged in this case, the logical part $\Gamma$ of the
precondition can soundly be reused in the postcondition.  In the rule
{\sc Q:Break}, where the linear control flow is broken, the
postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
taken from the break part of the judgement similarly as it is done in
the quantitative logic.  The rule {\sc Q:Return} exhibits a slight
difference to the quantitative logic.  While we have a function $\Z
\to \Assn$ to represent return values in the logic, we assume that a
special variable $\ret$ is part of the index set of $Q_R$.  The
substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
$q_{\ret y} = q'_{x y}$ and $q_{y \ret} = q'_{y x}$ for all $y$ and
$q_i'=q_i$ otherwise.

Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
  Q:Inc} for increments and decrements, which describe how the
potential is distributed after a size change of the variables.  The
rule {\sc Q:IncP} is for increments $x \gets x + y$ in which we can
deduce from the logical context $\Gamma$ that $y$ is not negative.
Similarly, {\sc Q:DecP} is for increments $x \gets x - y$ in which we
can deduce that $y \ge 0$.  Of course, we have symmetrical rules {\sc
  Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
The rules are equivalent in the case where $y=0$.  Finally, the {\sc
  Q:Inc} can be applied if we cannot deduce any information about $y$.
In the implementation, the rules for increments and decrements are
combined into one syntax directed rule where the specialized rules
take priority over {\sc Q:Inc}.

To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The other follow the
same idea and are symmetrical.  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  Remark that the rule insures this
only for $v \not\in \mathcal U$ because we know that $x \le v$
otherwise, and thus $|\inter v x| = 0$.

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the argument/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postconditions of the function
  specification is equalized to its matching potential in the callee's
  pre/postcondition.
\itemskipIn
\item
  The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local variables,
\itemskipIn
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, the rule takes care to split the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P + S$ and $Q' =
P' + S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function postcondition
$(\Gamma'_f, Q'_f)$.

The rule {\sc Q:Weak} is not syntax directed.  In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition.  We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the
analysis efficient.  The high-level idea of {\sc Q:Weak} is the same
as in the rule {\sc L:Weak} of the quantitative logic: If we have a
sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept
of \emph{more potential} is formalized by the relation $Q'
\succeq_\Gamma Q$ that is defined in the rule {\sc Relax}.  Here, $Q$
and $Q'$ are potential annotations and $\Gamma$ is a logical context.
The relax rule deals also with the important task of transferring
constant potential (represented by $q_0$) to interval sizes and vise
versa.  If we can deduce from the logical context that the interval
size $|\inter x y| \geq l$ is larger then a constant $l$ then we can
transfer potential $q_{xy} {\cdot} |\inter x y|$ form the interval to
constant potential $l{\cdot}q_{xy}$ and guarantee that we do not gain
potential.  Conversely, if $|\inter x y| \leq u$ for a constant $u$
then we can transfer constant potential $u{\cdot}q_{xy}$ to the
interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining
potential.

@@

We separate the search of a derivation into two steps.  As a first
step we go through the whole program and apply inductively the
inference rules of the automatic amortized analysis.  During this
process our tool uses symbolic names for the coefficients $(q_i)$.
Every time a linear constraint must be satisfied by these
coefficients, it is recorded using the symbolic names in a global
list.  When the list of all constraints that must be satisfied by the
coefficients is collected, we feed it to an off the shelf
LP-solver~\footnote{We currently use Coin-Or CLP.}.  If the solver
successfully finds a solution to the given problem, we know that a
derivation exists for the considered program.  We can then extract the
initial $Q$ from the solver and get a resource bound for the program.
To get a full derivation we simply extract the complete solution from
the solver, and apply it to the symbolic names $(q_i)$ of the
coefficients in the derivation.  If the LP-solver fails to find a
solution, an error is reported to the user.

@@

The translation of an assertion $(\Gamma, Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  We also need to
translate the assumptions in function contexts of the automatic
analysis.  This is done as follows.
%
\begin{align*}
\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
&\forall z \, \vec v \, v . (
  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs \mapsto \vec v]), \\
&\phantom{\forall z \, \vec v \, v . (}
  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret \mapsto v])
  )
\end{align*}
%
These definitions let us translate the judgement $J = B; R; P \vdash
S \dashv P'$ in the context $\Delta$ by
$$
  \tr(J) :=
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$
The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the analysis]
  If $J$ is a judgement derived with the rules of
  \pref{fig:auto}, then $\tr(J)$ is a quantitative
  Hoare triple derivable with the rules of
  \pref{fig:logic}.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and basically maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because we have to show that the preconditions of the rules
{\sc L:Weak} or {\sc L:Update}, respectively, are met.  \iffull{ For instance
we prove the following lemma to show the soundness of the translation
of {\sc Q:Weak}.

@@

We have experimentally evaluated the practicability of our automatic
amortized analysis with more than 30 challenging loop and recursion
patterns from open-source code and the
literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
examples together with the derived bounds is given in \pref{app:cat}.

@@

@@

\begin{figure*}
\centering
\begin{tabular}{r|c|c|c|c}
%
&
\begin{lstlisting}
while (y > x) {
  x++; tick(1); }
while (x > 2) {
  x -= 3; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (i > 100) {
  i--; tick(1) }
i += k+50;
while (i >= 0) {
  i--; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (x > 0) {
  x--;
  t=x, x=y, y=t;
  tick(1);
}
\end{lstlisting}
&
\begin{lstlisting}
assert(y >= 0);
while (x>y) {
  x -= y+1;
  for (z=y; z > 0; z--)
    tick(1);
  tick(1);
}
\end{lstlisting}
\\
%
\hline
& t08 (slightly modified t08a) & t19 & t30 & t15
%
\\
\hline
LOOPUS &
{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
{\small % $1 + \max(x, y)$ (???)
---
} &
---
\\
%
%
%
%
Rank &
{\small $2 {+} y {-} x (?)$ } &
{\small $54+k+i$ } &
--- &
{\small $2+2x-y$ }
\\
%
%
AAA 
  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
  & $|\inter 0 x| {+} |\inter 0 y|$
  & $|\inter 0 x|$
\end{tabular}
\caption{ Comparison of resource bounds derived by different tools on
  several examples with linear bounds.  AAA stands for our automatic
  amortized analysis for Clight.  The output of Rank has been manually
  simplified to fit the table.  }
\label{fig:compar}
\end{figure*}

\begin{figure}
\includegraphics[width=\linewidth]{fig/bound3d}
\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
  (blue lines) and the measured runtime cost (red crosses) for example
  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
\label{fig:3d}
\end{figure}

\begin{table}
\centering
\begin{tabular}{r|c|c|c|c|c}
& KoAT & Rank & LOOPUS & SPEED & AAA \\
\hline
\#bounds & 9 & 24 & 20 & 14 & 32
\\
\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
\\
\#best bounds & 0 & 0 & 11 & 14 & 29
\\
\#tested & 14 & 33 & 33 & 14 & 33
\end{tabular}
\caption{Comparison of our automatic amortized analysis with other
  automatic tools.  We have not been able to obtain a version of
  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
  does currently not work on C programs and we use the bounds that
  have been reported in the author's experimental evaluation.}
\label{tab:compar}
\end{table}

@@

@@

\bibliographystyle{abbrvnat}
\iffull{
\bibliography{lit}
}
{
\sectskip
\footnotesize
\bibliography{lit}
}

>>>>

%SHORT
\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
\captionsetup[figure]{format=myformat}
\captionsetup[table]{format=myformat}
\renewcommand{\captionlabelfont}{\bf}
%SHORT

@@

\iffull{We only show a small selections of the programs that we can
  handle automatically here.}{} \iffull{In \pref{app:cat}}{In the
  extended version~\cite{anon_extended}} is a list of more than $30$
classes of challenging programs that we can automatically analyze.
\pref{sec:exper} contains a more detailed comparison with other
tools\iffull{for automatic bound derivation.}{.}

@@

To find derivations in the quantitative Hoare logic automatically, we
have to focus on bounds that have a certain shape.  The general form
of \emph{potential functions} (or assertions) that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is again a simplified program state as introduced in \pref{sec:sem},
$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as\iffull{
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$}
{ $\Phi = \sum_{i \in I} q_i f_i$.}  We often write $xy$ to denote the
index $(x,y)$.  The family $(f_i)_{i\in I}$ is actually a basis (in
the linear-algebra sense) for potential functions.  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of \ifshort{non-negative}{} rational numbers where only
a finite number of elements are not zero. \iffull{ For soundness, we
  require that potential functions always give non-negative results.
  This is acheived by restricting coefficients in a quantitative
  annotation to be always non-negative.}{}

@@

In addition to the quantitative annotations our automatic amortized
analysis needs to maintain a minimal logical state to justify certain
operations on quantitative annotations.  For example when analyzing
the code $x \gets x + y$, it is helpful to know the sign of $y$ to
determine which intervals will increase or decrease.  The knowledge
needed by our rules can be inferred by local reasoning (i.e., in
basic blocks without recursion and loops) within usual theories
(e.g. Presburger arithmetic or bit vectors).

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
%
Our implementation uses conjunctions of linear inequalities as
logical state.  We never compute fixpoints and take $\top$
as pre- and postconditions for functions. It has proved to be
sufficient for the variety of examples we are interested in or found 
in the literature.

@@

\iffull{

@@

\and }{}
\Rule{Q:Return}

@@

\iffull{\and 

@@

}{}

@@

\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}

@@

\\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')

@@

\\\\ Q_1 \succeq_{\Gamma_1} Q_2 

@@

\end{mathpar}
\caption{\iffull{Inference rules}{Selected inference rules} of the quantitative analysis.}
\label{fig:auto}
\end{figure*}

@@

The inference system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The inference rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
These judgements correspond to the logic judgements $\Delta; B; R
\vdash \htriple P S Q$ where each assertion splits in two parts, a
logical part $\Gamma$ and a quantitative part $Q$.  That means that
$(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
$(\Gamma_R, Q_R)$ is the postcondition for return statements, and
$(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
simple Hoare triple.  We leave out the function context $\Delta$ in
the presentation to avoid cluttering notations.  Instead we assume a
fixed function context $\Delta$ that is implicit on all judgements.
Function contexts are treated exactly as in the logic.  % We call
% precondition the pair $(\Gamma, Q)$ and postcondition the pair
% $(\Gamma', Q')$.  
The correspondence between judgements of the
automation and the ones of the quantitative logic is made formal by
our soundness proof of the automatic amortized analysis at the end of
this section.

@@

We describe the automatic amortized analysis for a subset of
expressions of C light.  Assignments must have the form \code{x \gets
  y} or \code{x \gets x \pm y}.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This achieved by using a series of \emph{cost-free
  assignments} that do not result in additional cost in the
analysis.  Non-linear operations such as \code{x \gets z*y} are simply
handled by assigning zero potential to coefficients like $q_{xa}$ and
$q_{ax}$ that contain $x$ after the assignment.

\begin{figure*}
\centering
\begin{tabular}{r|c|c|c|c}
%
& t08 (slightly modified t08a) & t19 & t30 & t15\\
%
\hline
&
\begin{lstlisting}
while (y > x) {
  x++; tick(1); }
while (x > 2) {
  x -= 3; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (i > 100) {
  i--; tick(1) }
i += k+50;
while (i >= 0) {
  i--; tick(1); }
\end{lstlisting}
&
\begin{lstlisting}
while (x > 0) {
  x--;
  t=x, x=y, y=t;
  tick(1);
}
\end{lstlisting}
&
\begin{lstlisting}
assert(y >= 0);
while (x>y) {
  x -= y+1;
  for (z=y; z > 0; z--)
    tick(1);
  tick(1);
}
\end{lstlisting}
%
\\
\hline
LOOPUS &
{\small $\max(0, x{-}2) {+} 2\max(0, y {-} x)$} &
{\small $\max(0, \! i{-}10^2) \!+ \max(0, \! k {+} i {+} 51)$} &
{\small % $1 + \max(x, y)$ (???)
---
} &
---
\\
%
%
%
%
Rank &
{\small $2 {+} y {-} x (?)$ } &
{\small $54+k+i$ } &
--- &
{\small $2+2x-y$ }
\\
%
%
AAA 
  & {\small $1.33|[x,y]| {+} 0.33 |[0,x]|$ }
  & $50 {+} |\inter {-1} i| {+} |\inter 0 k|$ 
  & $|\inter 0 x| {+} |\inter 0 y|$
  & $|\inter 0 x|$
\end{tabular}
\vspace{.1cm}
\caption{ Comparison of resource bounds derived by different tools on
  several examples with linear bounds.  AAA stands for our automatic
  amortized analysis for Clight.  The output of Rank has been manually
  simplified to fit the table.  }
\label{fig:compar}
\end{figure*}

@@

Most of the rules correspond exactly to the respective rules of the
quantitative Hoare logic.  
\iffull{
  For example, the rule {\sc Q:Skip} simply reuses its postcondition
  as precondition.  This is sound since evaluating a $\code{skip}$
  statement never costs any resources and does not change the program
  state.  Similarly, the rule {\sc Q:Tick} enforces that the
  quantitative annotation of the precondition is at least $M_t(n)$ to
  pay for the resource consumption.  Since the program state is also
  unchanged in this case, the logical part $\Gamma$ of the
  precondition can soundly be reused in the postcondition.  In the
  rule {\sc Q:Break}, where the linear control flow is broken, the
  postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
  taken from the break part of the judgement similarly as it is done
  in the quantitative logic.
}
%
{ Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
  {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
  rules can be found in the extended version.
}
%
The rule {\sc Q:Return} exhibits a slight
difference to the quantitative logic.  While we have a function $\Z
\to \Assn$ to represent return values in the logic, we assume that a
special variable $\ret$ is part of the index set of $Q_R$.  The
substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
$q_{\ret y} = q'_{x y}$ and $q_{y \ret} = q'_{y x}$ for all $y$ and
$q_i'=q_i$ otherwise.

Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
  Q:Inc} for increments and decrements, which describe how the
potential is distributed after a size change of a variable.  The rule
{\sc Q:IncP} is for increments $x \gets x + y$ in which we can deduce
from the logical context $\Gamma$ that $y$ is not negative.
Similarly, {\sc Q:DecP} is for increments $x \gets x - y$ in which we
can deduce that $y \ge 0$.  Of course, we have symmetrical rules {\sc
  Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
The rules are equivalent in the case where $y=0$.  Finally, the rule
{\sc Q:Inc} can be applied if we cannot deduce any information about
$y$.  In the implementation, the rules for increments and decrements
are combined into one syntax directed rule where the specialized rules
take priority over {\sc Q:Inc}.

To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The other follow the
same idea and are symmetrical.  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ for the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  The rule insures this
only for $v \not\in \mathcal U$ because we know that $x \le v$
otherwise, and thus $|\inter v x| = 0$.

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the argument/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postconditions of the function
  specification is equalized to its matching potential in the callee's
  pre- and postcondition.
\itemskipIn
\item
  $\!$The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local\iffull{ variables,}{,}
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, {\sc Q:Call} splits the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P {+} S$ and $Q' =
P' {+} S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function's postcondition
$(\Gamma'_f, Q'_f)$.

The rule {\sc Q:Weak} is not syntax directed.  In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition.  \iffull{We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the
analysis efficient.}{}  The high-level idea of {\sc Q:Weak} is the same
as in the rule {\sc L:Weak} of the quantitative logic: If we have a
sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept
of \emph{more potential} is formalized by the relation $Q'
\succeq_\Gamma Q$ that is defined in the rule {\sc Relax}.  Here, $Q$
and $Q'$ are potential annotations and $\Gamma$ is a logical context.
The rule {\sc Q:Relax} deals also with the important task of transferring
constant potential (represented by $q_0$) to interval sizes and vise
versa.  If we can deduce from the logical context that the interval
size $|\inter x y| \geq l$ is larger then a constant $l$ then we can
transfer potential $q_{xy} {\cdot} |\inter x y|$ form the interval to
constant potential $l{\cdot}q_{xy}$ and guarantee that we do not gain
potential.  Conversely, if $|\inter x y| \leq u$ for a constant $u$
then we can transfer constant potential $u{\cdot}q_{xy}$ to the
interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining
potential.

@@

We separate the search of a derivation into two steps.  As a first
step we go through the whole program and apply inductively the
inference rules of the automatic amortized analysis.  During this
process our tool uses symbolic names for the rational coefficients
$(q_i)$ in the rules.  Every time a linear constraint must be
satisfied by these coefficients, it is recorded in a global list using
the symbolic names.  When then feed the collected constraints to an
off the shelf LP-solver~\footnote{We currently use Coin-Or's CLP.}.
If the solver successfully finds a solution, we know that a derivation
exists for the considered program.  We can then extract the initial
$Q$ from the solver and get a resource bound for the program.  To get
a full derivation we simply extract the complete solution from the
solver, and apply it to the symbolic names $(q_i)$ of the coefficients
in the derivation.  If the LP-solver fails to find a solution, an
error is reported to the user.

@@

The translation of an assertion $(\Gamma, Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  We also need to
translate the assumptions in function contexts of the automatic
analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) := \forall z \, \vec v \, v . $
%
$$
(
  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
  )
$$
%
These definitions let us translate the judgement $J = B; R; P \vdash
S \dashv P'$ in the context $\Delta$ by
$$
  \tr(J) :=
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$
The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the automatic analysis]
  If $J$ is a judgement derived with the rules of
  \pref{fig:auto}, then $\tr(J)$ is a quantitative
  Hoare triple derivable with the rules of
  \pref{fig:logic}.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and basically maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because we have to show that the preconditions of the rules
{\sc L:Weak} or {\sc L:Update}, respectively, are met.  \iffull{ For instance
we prove the following lemma to show the soundness of the translation
of {\sc Q:Weak}.

@@

We have experimentally evaluated the practicability of our automatic
amortized analysis with more than 30 challenging loop and recursion
patterns from open-source code and the
literature~\cite{GulwaniMC09,GulwaniJK09,GulwaniZ10}.  A full list of
examples \iffull{together with the derived bounds is given in
  \pref{app:cat}.}{is given in the extended
  version~\cite{anon_extended}.}

@@

\begin{table}[h]
\centering
\begin{tabular}{r|c|c|c|c|c}
& KoAT & Rank & LOOPUS & SPEED & AAA \\
\hline
\#bounds & 9 & 24 & 20 & 14 & 32
\\
\#lin.\ bounds & 9 & 21 & 20 & 14 & 32
\\
\#best bounds & 0 & 0 & 11 & 14 & 29
\\
\#tested & 14 & 33 & 33 & 14 & 33
\end{tabular}
\vspace{.1cm}
\caption{Comparison of our automatic amortized analysis with other
  automatic tools.  We have not been able to obtain a version of
  SPEED~\cite{GulwaniMC09} and just use the bounds that have been
  reported by the authors. Similarly, KoAT~\cite{BrockschmidtEFFG14}
  does currently not work on C programs and we use the bounds that
  have been reported in the author's experimental evaluation.}
\label{tab:compar}
\end{table}

@@

@@

\begin{figure}[t]
\includegraphics[width=\linewidth]{fig/bound3d}
\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
  (blue lines) and the measured runtime cost (red crosses) for example
  \emph{t08}. For $x\ge 0$ the derived bound is tight.}
\label{fig:3d}
\end{figure}

@@

\bibliographystyle{abbrvnat}
\iffull{
\bibliography{lit}
}
{
\sectskip
\scriptsize
\bibliography{lit}
}
